FraCaS: Temporal Analysis

### 2020-06-26 paper2020/paper.tex
spellchecking and abstract

@@

For the purpose of this paper, the important point is that the GF
syntax aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
Example \ref{ex:nobel} shows an example abstract syntax tree and its
realisation in English.

\paragraph{Coq}
Coq is an interactive theorem prover (proof assistant). It is based on
the calculus of coinductive constructions~\citep{werner_une_1994} We
do not use any coinduction (or even induction) in this paper, relying
on the pure lambda-calculus inner core of Coq.  Coq is a very powerful
reasoning engine that makes it fit for implementing natural language
semantics.  Coq supports, as already mentioned, dependent typing, but
also subtyping. Both concepts are instrumental in expressing NL
semantics\citet{TODO: chatzikyriakidis and luo}.  Besides, on a more
practical side, it works well for the the task of NLI, when the latter
is formalized as a theorem proving task: its many tactics mean that
many tasks in theorem proving are trivialized. In particular, all
problems of time-intervals inclusion, which occur in every temporal
problems, are solved with Coq's linear arithmetic tactic.

@@

In montagovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise the interpretation of 
verbs, so that it takes two
additional time parameters, one corresponding to the starting time
of the action and one corresponding to its stopping time ($(e × time × time) \to t$). For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). (We are thus
using a simple newtonian model of time, corresponding to a layman
intuition of a linear constant flow of time.)

@@

Even though we discuss a refinement to handle the past continuous
at the end of this section, the procedure to handle tense annotations
is as follows:
\begin{itemize}
\item If the tense is the past, and the temporal context is
  unspecified, then we locally quantify over a time interval
  $[t_0,t_1]$, such that $t_1 < now$, where $now$ is a logical
  constant representing the current timepoint. The temporal context
  then becomes this interval.
\item If the tense is the present and the temporal context is
  unspecified, then the temporal context becomes the simple
  $(now,now)$ interval.
\item If the temporal context is specified (for example due to the presence of an adverb or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$.
\end{itemize}

@@

A common theme in the testsuite is the reference to previous
occurences of an event:

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.
%
For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:

@@

Unicity of action holds only if the aspectual class of the proposition
$P$ is \emph{actitivty}~\citep{steedman_productions_2000} (which, for our
purposes, includes \emph{achievements} and \emph{accomplishments}).

(The difference between actitivty and accomplishments on the one hand and
achievement on the other hand is that for the latter, time intervals
can be assumed to be of nil duration. In reality, this is an
oversimplification as achievements are usually of short duration, but
not nil. However, this plays little role in our
analysis. As far as we can tell the FraCaS test suite does exercise temporal semantics to
such a level of precision.)

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because ``being in Paris'' is stative, we also get that
Smith was in Paris in any subinterval. Contrary to unicity of action,
subsumption is always valid.

\paragraph{Class-modifying adverbs}
It should be noted that some adverbs can locally disable the
application of $\constant {subsumption}$. For example, problem 299
features the sentence ``Smith lived in Birmingham for exactly a
year''.  Even though ``live'' is normally stative, one can no longer
apply subsumption in the context of ``exactly a year'' --- this can be
done by propagating another context flag in the montegovian semantics
(in addition to the temporal context).

@@

\fracasex{279}{
\item	Smith wrote a novel in 1991.
\hyp 	 Smith wrote it in 1992.
}
Indeed, following our account, the above (contradictory) inference problem is to be
interperted as

@@

We treat verbs in the progressive form as different semantically from
the non-progressive form. For example, ``John was writing a book'' is
encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, PROG\_write(John,book,t_1,t_2)$,
while ``John wrote a book'' is encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, write(John,book,t_1,t_2)$. This is
necessary, because $PROG\_write$ is subject to
$\constant{subsumption}$: if John is writing in the interval
$[t_1,t_2]$ then he is writing in any subinterval of
$[t_1,t_2]$. However, the same cannot be said of $write$: the
end-points of the interval indicate the time needed to complete the
achievement. (For example, ``John wrote a book in 1993'' neither
entails ``John wrote a book in January 1993'' nor ``John wrote a book
in December 1993''.) In fact, \emph{write}, in the non-progressive
from, is on the contrary subject to $\constant{unicity}$. Finally, we
also have $write(x,y,t_1,t_2) → PROG\_write(x,y,t_1,t_2)$. That is,
the achievement (or \emph{activity} in our terminology) variant implies the stative variant, in the
same interval.  Consequently we get the entailment from ``John wrote a
book'' to ``John was writing a book''.

@@

Our dynamic and temporal semantics gives the following interpretation
for \verb!s_279_1_p! implies \verb!s_279_3_h!.
{\small
\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}
}
In the above, one should remark the top-level quantification over the
novel (as explained in \cref{sec:our-semantics}), the quantification
over time intervals as individual timepoints, and the use of custom
operators for several constructions (\verb!FORALL!, \verb!Not!, \verb!IS_INTERVAL!,
\verb!appTime!). This use of custom operators is useful for several
generalisations (for example, we have quantifiers such as \verb!MOST! in
addition to \verb!FORALL!.)

Unfolding the definitions for these operators yield the following
proposition:
{\small
\begin{verbatim}
forall x : object,
novel_N x ->
(exists b c : Z,
   Date_19910101 <= b /\
   c <= Date_19911231 /\
   b <= c /\ write_V2 x SMITH b c) ->
(exists f g : Z,
   Date_19920101 <= f /\
   g <= Date_19921231 /\
   f <= g /\ write_V2 x SMITH f g) ->
False
\end{verbatim}
}This is very close to our idealised representation of the problem
\cref{eq:smith-novel-1}. One difference is the use of abstract Coq
integers for timepoints. Using a discrete time allows us to use
predifined Coq tactics. The discrete nature of integers does not
interfere with the reasoning.

@@

In the first of the two examples involving an achievement verb, the
inference is UNK, since there is no guarantee that the action is
non-instanteous. To the contrary, for accomplishment verbs, the
inference follows.

@@

\section*{Acknowledgements}

@@

>>>>

\begin{abstract}
  In this paper, we propose an implementation of temporal semantics
  which is suitable for inference problems. This implementation
  translates syntax trees to logical formulas, suitable for
  consumption by the Coq proof assistant. We support several phenomena
  including: temporal references, temporal adverbs, aspectual classes
  and progressives.  We apply these semantics to temporal problems of
  the FraCaS testsuite and obtain an accuracy of 71 percent.
\end{abstract}

@@

For the purpose of this paper, the important point is that the GF
syntax aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
Example \ref{ex:nobel} shows an example abstract syntax tree and its
realization in English.

\paragraph{Coq}
Coq is an interactive theorem prover (proof assistant). It is based on
the calculus of co-inductive constructions~\citep{werner_une_1994} We
do not use any co-induction (or even induction) in this paper, relying
on the pure lambda-calculus inner core of Coq.  Coq is a very powerful
reasoning engine that makes it fit for implementing natural language
semantics.  Coq supports, as already mentioned, dependent typing, but
also subtyping. Both concepts are instrumental in expressing NL
semantics\citet{TODO: chatzikyriakidis and luo}.  Besides, on a more
practical side, it works well for the the task of NLI, when the latter
is formalized as a theorem proving task: its many tactics mean that
many tasks in theorem proving are trivialized. In particular, all
problems of time-intervals inclusion, which occur in every temporal
problems, are solved with Coq's linear arithmetic tactic.

@@

In montagovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise the interpretation of 
verbs, so that it takes two
additional time parameters, one corresponding to the starting time
of the action and one corresponding to its stopping time ($(e × time × time) \to t$). For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). (We are thus
using a simple Newtonian model of time, corresponding to a layman
intuition of a linear constant flow of time.)

@@

Even though we discuss a refinement to handle the past continuous
at the end of this section, the procedure to handle tense annotations
is as follows:
\begin{itemize}
\item If the tense is the past, and the temporal context is
  unspecified, then we locally quantify over a time interval
  $[t_0,t_1]$, such that $t_1 < now$, where $now$ is a logical
  constant representing the current timepoint. The temporal context
  then becomes this interval.
\item If the tense is the present and the temporal context is
  unspecified, then the temporal context becomes the simple
  $(now,now)$ interval.
\item If the temporal context is specified (for example due to the presence of an adverb or an
  adverbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$.
\end{itemize}

@@

A common theme in the testsuite is the reference to previous
occurrences of an event:

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurrences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.
%
For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:

@@

Unicity of action holds only if the aspectual class of the proposition
$P$ is \emph{activity}~\citep{steedman_productions_2000} (which, for our
purposes, includes \emph{achievements} and \emph{accomplishments}).

(The difference between activity and accomplishments on the one hand and
achievement on the other hand is that for the latter, time intervals
can be assumed to be of nil duration. In reality, this is an
oversimplification as achievements are usually of short duration, but
not nil. However, this plays little role in our
analysis. As far as we can tell the FraCaS test suite does exercise temporal semantics to
such a level of precision.)

@@

Indeed, from \textbf{P3} we get that Smith was in Paris between May
5th and May 15th. Because ``being in Paris'' is stative, we also get that
Smith was in Paris in any sub-interval. Contrary to unicity of action,
subsumption is always valid.

\paragraph{Class-modifying adverbs}
It should be noted that some adverbs can locally disable the
application of $\constant {subsumption}$. For example, problem 299
features the sentence ``Smith lived in Birmingham for exactly a
year''.  Even though ``live'' is normally stative, one can no longer
apply subsumption in the context of ``exactly a year'' --- this can be
done by propagating another context flag in the montagovian semantics
(in addition to the temporal context).

@@

\fracasex{279}{
\item	Smith wrote a novel in 1991.
\hyp 	 Smith wrote it in 1992.
}
Indeed, following our account, the above (contradictory) inference problem is to be
interpreted as

@@

We treat verbs in the progressive form as different semantically from
the non-progressive form. For example, ``John was writing a book'' is
encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, PROG\_write(John,book,t_1,t_2)$,
while ``John wrote a book'' is encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, write(John,book,t_1,t_2)$. This is
necessary, because $PROG\_write$ is subject to
$\constant{subsumption}$: if John is writing in the interval
$[t_1,t_2]$ then he is writing in any sub-interval of
$[t_1,t_2]$. However, the same cannot be said of $write$: the
end-points of the interval indicate the time needed to complete the
achievement. (For example, ``John wrote a book in 1993'' neither
entails ``John wrote a book in January 1993'' nor ``John wrote a book
in December 1993''.) In fact, \emph{write}, in the non-progressive
from, is on the contrary subject to $\constant{unicity}$. Finally, we
also have $write(x,y,t_1,t_2) → PROG\_write(x,y,t_1,t_2)$. That is,
the achievement (or \emph{activity} in our terminology) variant implies the stative variant, in the
same interval.  Consequently we get the entailment from ``John wrote a
book'' to ``John was writing a book''.

@@

Our dynamic and temporal semantics gives the following interpretation
for \verb!s_279_1_p! implies \verb!s_279_3_h!.
{\small
\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}
}
In the above, one should remark the top-level quantification over the
novel (as explained in \cref{sec:our-semantics}), the quantification
over time intervals as individual timepoints, and the use of custom
operators for several constructions (\verb!FORALL!, \verb!Not!, \verb!IS_INTERVAL!,
\verb!appTime!). This use of custom operators is useful for several
generalizations (for example, we have quantifiers such as \verb!MOST! in
addition to \verb!FORALL!.)

Unfolding the definitions for these operators yield the following
proposition:
{\small
\begin{verbatim}
forall x : object,
novel_N x ->
(exists b c : Z,
   Date_19910101 <= b /\
   c <= Date_19911231 /\
   b <= c /\ write_V2 x SMITH b c) ->
(exists f g : Z,
   Date_19920101 <= f /\
   g <= Date_19921231 /\
   f <= g /\ write_V2 x SMITH f g) ->
False
\end{verbatim}
}This is very close to our idealized representation of the problem
\cref{eq:smith-novel-1}. One difference is the use of abstract Coq
integers for timepoints. Using a discrete time allows us to use
predefined Coq tactics. The discrete nature of integers does not
interfere with the reasoning.

@@

In the first of the two examples involving an achievement verb, the
inference is UNK, since there is no guarantee that the action is
non-instantaneous. To the contrary, for accomplishment verbs, the
inference follows.

@@

\section*{Acknowledgments}

@@

% LocalWords:  dowty steedman higginbotham fernando NLI FraCaS GF une
% LocalWords:  aspectual dataset anaphora bernardy grammaticality NL
% LocalWords:  infelicitousness ranta Ljunglof montague english TODO
% LocalWords:  prover chatzikyriakidis Coq's anonymized monadic unger
% LocalWords:  charlow anaphoroi stative activites timespan timespans
% LocalWords:  useCl scandinavian nobel FraCas timepoint Itel unicity
% LocalWords:  Statives contrario th Un lexemes PROG lexicalized lia
% LocalWords:  postfix timepoints FORALL appTime writeUnique FC CCG
% LocalWords:  FraCoq Mineshima Langpro abzianidze premiss UNK Centre
